---
title: "Primer of Ecology using R"
author: "Hank Stevens"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
delete_merged_file: true
output:
    bookdown::gitbook
    # bookdown::tufte_book:
    #  includes:
    #    in_header:preamble.tex
documentclass: book
bibliography: [/Users/stevenmh/MyDrive/library.bib, book.bib, packages.bib]
biblio-style: plain
link-citations: yes
github-repo: HankStevens/Primer-of-Ecology
output_dir: "docs"
---

```{r include=FALSE}
knitr::opts_chunk$set(echo = TRUE, results='markup', warning=FALSE, message=FALSE, fig.path = "figs/", knitr.graphics.auto_pdf = TRUE,
                      message=FALSE, out.width="90%")

library(bipartite)
library(cowplot)
library(data.table)
library(diagram)
library(DiagrammeR)
library(DiagrammeRsvg)
library(igraph)
library(kableExtra)
library(knitr)
library(latticeExtra)
library(lavaan) #sem for bipartite stability hypotheses
library(magrittr)
library(phaseR)
library(primer)
library(reshape2)
library(rARPACK) # needed for eigs in bip_stability
library(rsvg)
library(scatterplot3d)
library(semPlot)
library(tidyverse)
library(tufte)
library(untb)
library(vegan)
library(xtable)

theme_set(theme_minimal() + 
theme(panel.grid.major = element_line(colour = "darkgrey"),
      panel.grid.minor = element_line(colour = "darkgrey")) ) 
theme(panel.grid.major = element_line(colour = "darkgrey"))
cbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", 
               "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
```

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```


# Preface {-}

This is a draft in progress, based on my book published with Springer, "A Primer of Ecology With R", part of the Use R! series. There will eventually be chapters on multiple basins of attraction, food webs, networks, and within-trophic-level community structure and species diversity.

This book is dedicated to Dr. Zac Long, whose desire to publish with Robert Peters and Donald Strong has inspired me to continue to strive for my goals.

\paragraph{Goals and audience}
In spite of the presumptuous title, my goals for this book are modest. I wrote it as 


* the manual I wish I had in graduate school, and
* a primer for our graduate course in Population and Community Ecology at Miami University.^[Miami University is located in the Miami River valley in Oxford, Ohio, USA; the region is home to the Myaamia tribe that dwelled here prior to European occupation.]

------------------
To install the latest version of the `primer` R package, install it directly from GitHub. To do this, install the `devtools` package. To successfully do that on a Windows operating system, I think you need to install Rtools:

`https://cran.r-project.org/bin/windows/Rtools/`


Once you have done all that, use this code:

```
devtools::install_github("HankStevens/primer")
```

------------------

