---
title: "Primer of Ecology using R"
author: "Hank Stevens"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
delete_merged_file: true
output:
    bookdown::bs4_book:
       includes:
         in_header: analytics_tag_preface.html
documentclass: book
bibliography: [/Users/stevenmh/library.bib, book.bib, packages.bib]
biblio-style: plain
link-citations: yes
github-repo: HankStevens/Primer-of-Ecology
output_dir: "docs"
clean: FALSE
description: "A primer of R and of population and community ecology."
---

```{r include=FALSE}
knitr::opts_chunk$set(echo = TRUE, results='markup', warning=FALSE, message=FALSE, fig.path = "figs/", knitr.graphics.auto_pdf = TRUE,
                      message=FALSE, out.width="90%")

library(bipartite)
library(cowplot)
library(data.table)
library(diagram)
library(DiagrammeR)
library(DiagrammeRsvg)
library(igraph)
library(kableExtra)
library(knitr)
library(latticeExtra)
library(lavaan) #sem for bipartite stability hypotheses
library(magrittr)
library(phaseR)
library(primer)
library(reshape2)
library(rARPACK) # needed for eigs in bip_stability
library(rsvg)
library(scatterplot3d)
library(semPlot)
library(tidyverse)
library(tufte)
library(untb)
library(vegan)
library(xtable)

theme_set(theme_minimal() + 
theme(panel.grid.major = element_line(colour = "darkgrey"),
      panel.grid.minor = element_line(colour = "darkgrey")) ) 
theme(panel.grid.major = element_line(colour = "darkgrey"))
cbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", 
               "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
```

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```


# Preface {-}

NEW IN THIS EDITION OF THE E-BOOK (20 August 2021)

* Lotka-Volterra versions of direct competition AND mutualism are together in one chapter, followed by consumer-resource interactions, including predator-prey, and resource-based competition and mutualism.
* a chapter on consumer-resource competition and mutualism that follows chapters covering two-species consumer resource interactions (mostly predator-prey). 
* If it has been > 1 y since looking at the `primer` package, you might want to peruse the index - I added a few more functions including ones for 1/f noise, spectral mimicry, and the Allee effect.

------
This is based on my book published with Springer, "A Primer of Ecology With R", part of the Use R! series.

The `primer` package on CRAN (v. 1.2.0) should suffice for this edition. You may also install it directly from GitHub. To do this, install the `devtools` package. To successfully do that on a Windows operating system, I think you need to install Rtools:

`https://cran.r-project.org/bin/windows/Rtools/`


Once you have done all that, use this code:

```
devtools::install_github("HankStevens/primer")
```

In spite of the presumptuous title, my goals for this book are modest. I wrote it as 

* the manual I wish I had in graduate school, and
* a primer for our graduate course in Population and Community Ecology at Miami University.^[Miami University is located in the Miami River valley in Oxford, Ohio, USA; the region is home to the Myaamia tribe that dwelled here prior to European occupation.]

**Acknowledgements**

I am grateful for the generosity of early reviewers and readers, each of whom has contributed much to the quality of this work: Jeremy Ash, Tom Crist, David Gorchov, Raphael Herrera-Herrera, Thomas Petzoldt, James Vonesh, as well as several anonymous reviewers, and the students of our Population and Community Ecology class over the past 15 years. I am also grateful for the many conversations and emails shared with four wonderful mathematicians and theoreticians: Jayanth Banavar, Ben Bolker, Stephen Ellner, and Steve Wright --- I never have a conversation with these people without learning something. I have been particularly fortunate to have team-taught Population and Community Ecology at Miami University with two wonderful scientists and educators, Davd Gorchov and Thomas Crist. Only with this experience, of working closely with these colleagues, have I been able to attempt this book. It should go without saying, but I will emphasis, that the mistakes in this book are mine, and there would be many more but for the sharp eyes and insightful minds of many other people.

I am also deeply indebted to the R Core Development Team for creating, maintaining and pushing forward the R programming language and environment. Like the air I breathe, I cannot imagine my (professional) life without it. I am grateful to all of the developers of packages that have allowed literate programming with Rmarkdown and LaTeX, using RStudio and GitHub. 

I am indebted to Rachel Collins for her size-structured coneflower data, to Scott Meiners and his colleagues for their generous sharing of data, metadata, and statistical summaries from the Buell-Small Succession Study (http://www.ecostudies.org/bss/). I would like to thank Stephen Ellner for Ross's Bombay death data and for R code and insight over the past few years.  I am also indebted to Tom Crist and his colleagues for sharing some of their moth data (work supported by The Nature Conservancy Ecosystem Research Program
NSF DEB-0235369).

Last, and most importantly, I would like to thank those of you in my life who make living worthwhile.

Martin Henry Hoffman Stevens

Oxford, OH, USA, Earth

Today, Current Era



