---
title: "Primer of Ecology with R"
author: "Hank Stevens"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
delete_merged_file: true
output:
    # bookdown::tufte_book2:
    #   includes:
    #     in_header: preamble.tex
    bookdown::gitbook
documentclass: book
bibliography: [/Volumes/GoogleDrive/My Drive/library.bib, book.bib, packages.bib]
biblio-style: plain
link-citations: yes
github-repo: rstudio/bookdown-demo
description: "A Primer of Ecology"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, results='markup', warning=FALSE,
                      fig.path = "figs/", knitr.graphics.auto_pdf = TRUE,
                      message=FALSE, out.width="90%")
library(primer)
library(deSolve)
library(latticeExtra)
library(tidyverse)
theme_set(theme_minimal() + 
theme(panel.grid.major = element_line(colour = "darkgrey"),
      panel.grid.minor = element_line(colour = "darkgrey")) ) 
theme(panel.grid.major = element_line(colour = "darkgrey"))
cbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", 
               "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
library(diagram)
library(DiagrammeR)
library(DiagrammeRsvg)
library(knitr)
library(tufte)
library(reshape2)
library(data.table)
library(magrittr)
library(kableExtra)
library(rsvg)
library(igraph)
library(phaseR)

```

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```


# Preface {-}

\paragraph{Goals and audience}
In spite of the presumptuous title, my goals for this book are modest. I wrote it as 
\begin{itemize}
  \item the manual I wish I had in graduate school, and
\item a primer for our graduate course in Population and Community Ecology at Miami University\footnote{Miami University is located in the Miami River valley in Oxford, Ohio, USA; the region is home to the Myaamia tribe that dwelled here prior to European occupation.}
\end{itemize}


