# Host-parasitoid~relations

A parasitoid is a ghastly thing. These animals, frequently small wasps, characteristically lay an egg on their hosts, often a caterpillar, the young hatch out, and then slowly consume the juicy bits of the host. Neither the adult wasp nor their larvae immediately kill their host. Many eggs can be laid on a host, and many larvae can live (at least briefly) in the host. Ultimately, however, the host is gradually consumed and one or a few larvae per host metamorphosizes into an adult. Thus parasitoid larvae always consume and kill their hapless host. In this sense, their natural history is intermediate between that of predator and parasite---they are parasite-like, or a parasitoid. Thank the gods for parasitoids, as they often help suppress other animals that we like even less, such as herbivores that eat our crops. There are even parasioids of parasitoids, or hyperparasitoids. And they are among the most diverse taxonomic groups of all animals.

There are a variety of characteristics about host--parasitoid relations that might make them different from the kind of predator--prey relations that we have been thinking about up until now. In particular, the life cycles of host and prey are so intimately connected that there is a one-to-one correspondence between dead hosts and the number of parasitoids in the next generation. If we know how many hosts are killed, then we know approximately how many parasitoids are born.

## Independent and random attacks

In keeping with convention of parasitoid models, let us describe dynamics of hosts and their parasitoid enemies with a discrete time model [@May:1978fk]. This makes sense, as these organisms, typically insects, frequently complete their life
cycles together and within a year. Let us pretend that the hosts, $H$,
grow geometrically in the absence of parasitoids, such that $H_{t+1} =
RH_t$. If we assume that some individuals are attacked and killed by
parasitoids, $H_a$, then this rate becomes 
\begin{equation}
 (\#eq:12)
  H_{t+1} = R\left( H_t - H_a \right)
\end{equation}
Further, we assume a couple things about parasitoid natural history.
First, we pretend the parasitoids *search randomly and
independently* for prey, and are able to search area, $a$ ("area of discovery"), in their life time. The total number of *attack events* then is $E_t=aH_tP_t$: adults may lay an egg on any host it encounters, but this event does not kill the host. Therefore, hosts can receive more than one egg, thereby making the number of attacked hosts lower than the number of attack events, $E_t$. Each attack occurs as a random event, and for now we assume that each attack is independent of all others.

For many species of parasitoids, only one adult can emerge from a single host, regardless of how many eggs were laid on that host. Therefore, the number of emerging adult parasitoids, $P_{t+1}$, is simply the number of hosts that were attacked at all, whether the host received one egg or many. Therefore, the number of parasitoids at time $t+1$ is merely the number of hosts that were attacked, and we can represent this as 
\begin{equation}
  (\#eq:13)
  P_{t+1} = H_a.
\end{equation}
The assumption of one dead host = one new parasitoid can be easily relaxed if we introduce a constant $q$, such that $P_{t+1} = qH_a$. 

@Nicholson:1935aa took advantage of a well known discrete probability distribution, the *Poisson distribution*^[This distribution was discovered by a French mathematician (Sim\'{e}on-Denis Poisson (1781--1840), so we pronounce the name "pwah-sohn," like the "fois" of fois gras, and the "sohn" with a quiet "n," like, well, like the French would say it. It is the series
\begin{equation*}
  \frac{1}{e^\mu},  \frac{\mu}{1!e^\mu}, \frac{\mu^2}{2!e^\mu}, \cdots \frac{\mu^r}{r!e^\mu}
\end{equation*}
representing the probabilities of occurrence of 0, 1, 2, \ldots $r$.] to create a  simple discrete analytical model of host--parasitoid dynamics. 
\begin{align}
  H_{t+1} &= R H_te^{-aP_t}  (\#eq:para1)\\
  P_{t+1} &= H_t \left(1-e^{-aP_t}\right)(\#eq:para2)
\end{align}
Here $aP_t$ is  the mean number of attacks per larva ($\lambda$ of the Poisson distribution) and results from $P_t$ parasitoids searching a particular area, $a$. Parameter $a$ is known as the "area of discovery," the area searched during a parasitoid's lifetime --- more area searched (i.e., larger $a$) means more hosts killed, and fewer to survive and contribute to the next generation. The Poisson distribution tells us that if attacks are random and independent, then $e^{-aP_t}$ is the probability of a host not being attacked. Therefore, $H_te^{-aP_t}$ is the expected number of hosts which are not attacked and survive to reproduce. 

We can find the equilibrium for this discrete model. Recall that, in models of continuous differential equations, this meant letting $\D N/\D t =0$. In the discrete case, it means letting  $N_{t+1}-N_t=0$. The equilibrium, $N^*$, is the value of $N$ when this is true, so $N^*=N_{t+1}=N_t$. We use this to find the equilibrium of the parasitoids, starting with \@ref(eq:para1),
\begin{align}
  H^*&=RH^*e^{-aP_t}\notag\\
  1 &=Re^{-aP_t} \notag\\
  P^*&=P_t=\frac{\log R}{a}
\end{align}
If we recognize that in \@ref(eq:para2), $P_{t+1}=H_t-H_{t+1}/R$, and that $H^*=H_t=H_{t+1}$, then we find the host equilibrium
\begin{align}
  H^*&=P^*\frac{R}{R-1}.
\end{align}

### Simulating simple host-parasitoid dynamics
Here we motivate the Nicholson-Bailey model with a simulation of these dynamics. It may provide a more intuitive understanding than simply throwing a probability distribution at the problem. In the end, we wind up at the same place, and this should be reassuring.

The Nicholson-Bailey model  assumes that parasitoids search a fixed area, and each attacks hosts randomly and independently. We set the duration of the time series, the model parameters, and create an empty data frame to hold the results.
```{r}
time <- 20; R <- 3; a <- .005
HPs <- data.frame(Hs <- numeric(time), Ps <- numeric(time))
```
Next we calculate the equilibrium, and use a nearby point for the initial abundances.
```{r}
Pst <- log(R)/a; Hst <- Pst*R/(R-1) 
HPs[1, ] <- c(Hst+1, Pst)
```
We then project the dynamics, one generation at a time, for each time step.
```{r}
for(t in 1:(time-1))  HPs[t+1,]  <- {H <- R * HPs[t,1] * exp( -a * HPs[t,2] )
                                 P <- HPs[t,1]*(1-exp(-a*HPs[t,2]))
                               c(H,P) }
```

Last we plot the trajectory, placing a point at the (unstable) equilibrium, and using arrows to highlight the direction and magnitude of the increment of change per time step.
```{r fig.cap="*Dynamics around the unstable equilibrium of the Nicholson-Bailey host--parasitoid model ($R=3$, $a=0.005$). Arrows indicate a single time step; the point is the equilibrium at the center.*"}
plot(HPs[,1], HPs[,2], type="n", xlab="H", ylab="P"); points(Hst, Pst);
arrows(HPs[-time,1], HPs[-time,2], HPs[-1,1], HPs[-1,2], length=.05)
```

#### Simulating Random Attacks
This model makes strong assumptions about the proportion of hosts that escape attacks, that depends on parasitoid attacks being random and independent. Therefore let us simulate parasitoid attacks that are random and independent and compare those data to field data. Let's start with some field data collected by Mark Hassell [@May:1978fk], which are the number of parasitoid larvae per host, either 0, 1, 2, 3, or 4 larvae.
```{r}
totals <- c(1066, 176, 48, 8, 5)
```
Here, 1066 hosts have no larvae. We can recreate the data set, where we have one observation per host: the number of larvae in that host.
```{r}
obs <- rep(0:4, totals)
```
To simulate random attacks, let's use the same number of hosts, and the same number of attacks, and let the computer attack hosts at random. We calculate the total number of hosts, $H$, and the total and mean number of attacks experienced by hosts.
```{r}
H <- sum( totals )
No.attacks <- sum(obs)
mean.attacks <- mean(obs)
```
Next, the predators attack (or 'sample') the hosts at random and with
equal probability. To code this, we identify the hosts by numbering
them 1--$H$. We then attack these hosts independently, that is, we ```replace``` each attacked host back into the pool of possible prey. 
```{r}
attack.events <- sample(x=1:H, size=No.attacks, replace=TRUE)
```
We then count the number times different hosts were attacked.
```{r}
N.attacks.per.host <- table(attack.events)
```
and then find count the number of hosts that were attacked once, twice, or more.
```{r}
(tmp <- table( N.attacks.per.host ) )
```
We see, for instance, that \Sexpr{tmp[2]} hosts were attacked twice. This also allows us to know how many hosts were \emph{not} attacked,
```{r}
not.att <- H - sum(tmp)
```
Let's compare the observed data to the simulated data.
```{r}
obs.sim <- rep(0:max(N.attacks.per.host), c(not.att, tmp))
table(obs)
table(obs.sim)
```
*There seem to be fewer unattacked hosts and more attacked hosts in the random-attack simulation than in the observed data*. Is this simply a consequence of the observed values being one of many random possibilities? Or is it a systematic difference? How do we check this?

One way to check whether the observed data differ from our model of random and independent attacks is to simulate many such observations, and compare the observed data (```rtotals[1]``` unattacked hosts) to the distribution of the simulated results. Here we do this by performing $n$ simulations of the same steps we used above.
```{r}
n <- 1000
unatt.sim<- sapply(1:n, function(j) { ## Repeat n times...
  ## simulate the same number of attacks.
  host.sim.att <- sample(x=1:H, size=No.attacks, replace=TRUE)
  ## subtract the number of attacked hosts from the total in the host population
  ## The number of attacks on each attacked host
  attacks.per <- table(host.sim.att)
  ## the number of attacked hosts 
  n.attd <- length( attacks.per)
  ## the number not attacked
  H - n.attd
                         } )
``` 
Next we just make a histogram (Fig. \@ref(fig:PoisSim)) of the number of unattacked hosts, adding a dotted line to show where the true data lie, and a dashed line for the prediction, under the assumption of random and independent attacks, based on the Poisson distribution.
```{r PoisSim, fig.cap="*Histogram of simulated host populations, attacked at a rate of ```r round(mean.attacks,2)``` mean attacks per host, assuming a attacks on hosts are random and independent of each other.*"}
hist(unatt.sim, xlab="Simulated # of Unattacked Hosts", 
     prob=TRUE, xlim=c(1000, 1070) )
abline(v=1066, lty=3)
abline(v=exp(-mean.attacks), lty=2)
```

Our simulation results (Fig. \@ref(fig:PoisSim)) indicate that the observed data include far more unattacked hosts than expected by chance.

Another, quicker way to evaluate the assumption of independent and random attacks is to compare the ratio of the variance of the observed larvae per host to the mean. If the data follow a Poisson distribution, this ratio is expected to equal to one.
```{r}
(I <- var(obs)/mean(obs))
```
This ratio is greater than one, but we do not know if this could be due to chance. We can test it, because under the null hypothesis, we expect that the product of the ratio and the number of hosts, $H$, follows a $\chi^2$ distribution, with $H-1$ degrees of freedom. We can ask how likely this ratio, or a more extreme ratio, is, if attacks are random and independent. We compare it to the cumulative probability density function for the $\chi^2$ distribution. 
```{}
1-pchisq(I*H, df=H-1)
```
We find that, given this large number of observations (1303 hosts), it is exceedingly unlikely to observe a ratio this large or larger. It is nice that this agrees with our simulation! We feel more confident when a parametric test agrees with a simulation of our ideas; perhaps both are not necessary, but working through both helps us to understand what we are doing.

## Aggregation leads to coexistence
The above model (\@ref(eq:para1), \@ref(eq:para2)) predicts that the parasitoid or host becomes extinct. That is potentially useful, because it suggests that we are missing something, and that is a feature of models that we like: they allow us to rigorously evaluate our assumptions and falsify them. All scientific ideas ultimately have to be described in a manner that could allow us to, in principle, falsify them. If we have some confidence in the model above, this falsification might lead us to ask, "what is the model missing?"

Perhaps what we are missing is some important parasitoid natural history. Parasitoids tend to aggregate on particular hosts or on local populations of the hosts. Some hosts are more likely to get attacked than others, resulting in more unattacked hosts and more hosts receiving multiple attacks than predicted under random and independent attacks. This may be related to their searching behavior, or to spatial distributions of hosts in the landscape. When the parasitoids aggregate, they tend to leave the low density host populations unparasitized. As a result, these low density populations act as *refuges*.

Perhaps we can kill two birds with one stone, and fix both problems with one step. That is what @May:1978fk and others have done, by assuming that parasitoids *aggregate*. 

May proposed the following logic for one reason we might observe more unattacked hosts than expected. Imagine that the distribution of parasitoids *among* patches in the landscape can be described with one probability distribution, with some particular mean and variance in the number of parasitoids per patch. Imagine next that their attacks on hosts *within* each patch are random and independent and thus described with the Poisson distribution. This will result in a compound distribution of attacks --- a combination of the among-patch, and within-patch distributions.^[This distribution is actually a mixture of Poisson distributions. The Poisson distribution is determined entirely by its mean, $\lambda$. If we let $\lambda$ itself be a random variable drawn from the Gamma distribution, then we arrive at the negative binomial distribution.] If we examine the distribution of larvae per host, for all hosts in the landscape, we will find a higher proportion of unattacked hosts, and a higher proportion of hosts that are attacked many times. 

The *negative binomial distribution* can describe aggregatred data, such as the number of larvae per host. Such data are counts greater than or equal to zero, in which the variance is greater than the mean. Thus, the negative binomial distribution can describe greater aggregation than the Poisson distribution (where $\mu=\sigma^2$), and thereby describe nature somewhat more accurately. May suggested that while the true distribution of larvae in hosts, in any particular case, was unlikely to truly be a negative binomial, it was nonetheless a useful approximation.

Ecologists frequently use the negative binomial distribution for data where the variance is greater than the mean. There are different derivations of the distribution, and therefore different parameterizations [@Bolker:2008rr]. In ecology, we typically use the mean, $\mu$, and the *overdispersion parameter* $k$. The variance, $\sigma^2$, is a function of these, where $\sigma^2 = \mu + \mu^2/k$; by overdispersion we mean that $\sigma^2 > \mu$. Large values of $k$ ($k>10$) indicate random and independent events rather than aggregation, and the distribution becomes indistinguishable from the Poisson. Small values ($k<2$) indicate aggregation.^[Although $k$ is often referred to as a "clumping" or "aggregation" parameter, we might think of it as a "randomness" parameter, because larger values result in more random, Poisson-like distributions.] Fig. \@ref(fig:negbin) shows examples. R has the negative binomial distribution built in (`nbinom`), but does not use $k$ as one of its arguments; rather, `size` $=k$. Here we generate a graph showing the distribution with different values of $k$.
```{r eval=FALSE, echo=FALSE}
getk <- function(CV2, mu){mu/(mu*CV2-1)}
getk(7.33, -.24)
getk(7.77, -.46)
getk(2, -.025)
getk(.1,.395)
```
```{r negbin, fig.cap="*The negative binomial distribution, where the dispersion parameter $k$ controls variance or breadth the distribution. For a given mean, smaller $k$ causes a greater variance, and results in a higher proportion of zeroes. These $k$ are representative of values from @Pacala1990.*"}
nb.dat <- cbind(
       Random = dnbinom(0:5, mu=.5, size=10^10),
      "k=10"=dnbinom(0:5, mu=.5, size=10),
      "k=1"=dnbinom(0:5, mu=.5, size=1),
      "k=0.01"=dnbinom(0:5, mu=.5, size=.01)
               )
matplot(0:5, nb.dat , type="b", pch=1:4, col=1, ylim=c(0,1),
        xlab="Attacks per Hosts", ylab="Probability") 
legend("topright", rev(colnames(nb.dat)), pch=4:1, lty=4:1, bty='n')
mtext(quote(mu== 0.5), padj=2)
```

The proportion of unattacked hosts expected under the negative binomial is $(1 + aP_t/k)^{-k}$. Therefore, we can write the analytical expressions for the population dynamics that are very similar to those of the Nicholson-Bailey model, but using the negative binomial distribution.
\begin{align}
  H_{t+1} &= RH_t\left(1+\frac{aP_t}{k} \right)^{-k}  \label{eq:may14a}\\
  P_{t+1} &=  H_t - H_t\left(1+\frac{aP_t}{k} \right)^{-k}  \label{eq:may14b}
\end{align}
@May:1978fk referred to this as a phenomenlogical model (as opposed to mechanistic) because the negative binomial merely approximates the true, perhaps compound, distribution.

### Equilibria for a discrete-time model
The equilibria of the host and parasitoid populations in May's model (eqs. \@ref(eq:may14a), \@ref(eq:may14b)) are  derived simply, once we decide upon how to describe the condition for an equilibrium (a.k.a. steady state). In models of differential equations, this meant letting $d N/d t =0$. In the discrete case it means letting  $N_{t+1}-N_t=0$. The equilibrium is the value of $N$ when this is true, so $N^*=N_{t+1}=N_t$. Following this train of thought, we have $H^*=H_{t+1} = H_t $ and $P^*=P_{t+1} = P_t$. 

To solve for equilbria, we begin with the expression for hosts (\@ref(eq:may14a)), and solve for equilibrium parasitoid density. In this case, we can divide \@ref(eq:may14a) both sides by $H^*$.
\begin{align}
  1 &= R\left(1+\frac{aP_t}{k}\right)^{-k}\notag\\
  R^{-1} &=  \left(1+\frac{aP_t}{k}\right)^{-k}\notag\\
  R^{1/k} &= 1 + \frac{aP_t}{k}\notag\\
  P^* &= \frac{k}{a} \left(R^{1/k} - 1 \right)\label{eq:16}
\end{align}
Given this, what causes increases in $P^*$? Certainly decreasing $a$ leads to increases in $P^*$. If a parasitoid population has a smaller $a$ (i.e. smaller area of discovery), this means that they require less space, and can thereby achieve higher density. Increasing $k$ means less aggregated attack events, which increases the probability that a parasitoid larvae is alone on a host, and this increases parasitoid density, but only up to  a limit.

If we want $H^*$ as well, we can solve for that too. One of many ways is to realize that, \@ref(eq:may14b) can be rearranged to show
\begin{align*}
  P_{t+1} &= H_t - \frac{H_{t+1}}{R}
\end{align*}
and given contant $H$ and $P$, 
\begin{align}
  (\#eq:18)
  H^* &= P^* \left(\frac{R}{R-1} \right)
\end{align}
where $P^*$ is \@ref(eq:may14a). 


## Stability of host--parasitoid dynamics
As with differential equation models, we can analyze the stability in the immediate vicinity of equilibria in discrete time models using eigenanalysis of the Jacobian matrix. Although analogous, there are some important differences  between continuous and discrete models.

* The Jacobian matrix is comprised of partial derivatives of growth increments (e.g., $\Delta N = N_{t+1}-N_t$), rather than of time derivatives (e.g., $\dot{N}$).
* Interpretation of eigenvalues of the Jacobian reflects the difference between finite rate of increase ($\lambda$) and the intrinsic rate of increase ($r$); e.g., for discrete models, $\lambda = 1$ indicates zero growth.
* Populations with discrete generations can grow too quickly to be stable (e.g., chaos in discrete logistic growth).


In the discrete time case, the Jacobian matrix is the set of partial derivatives of the discrete growth increment rather than of the time derivatives used for continuous growth^[Note $dN/dt$ is also an increment --- it is the increment $\Delta N$ as $\Delta t \to 0$, whereas in the discrete case, $\Delta t = 1$.]. The growth increments are the increments of change over an entire time step, or $\Delta H=H_{t+1} - H_t$ and $\Delta P=P_{t+1} - P_t$. Taking those increments from \@ref(eq:may14a), \@ref(q:may14b), the Jacobian matrix of partial differential equations is 
\begin{align}
  (\#eq:jachp)
\left(
\begin {array}{cc}
\frac{\partial \Delta H}{\partial H}&\frac{\partial \Delta H}{\partial P}\\
\frac{\partial \Delta P}{\partial H}&\frac{\partial \Delta P}{\partial P}
\end {array}
\right) =
\left(
\begin {array}{cc}
R\left(1+aP\right)^{-k}-1\quad &-akHR\left(1+aP\right)^{-\left(k+1\right)} \\
1-\left(1+aP\right)^{-k}& akH\left(1+aP\right)^{-\left(k+1\right)}
\end {array}
\right) 
\end{align}
@ 
To analyze this at the equilibrium, we substitute $H^*,\,P^*$ for $H,\,P$, and perform eigenanalysis. The resulting eigenvalues, $\lambda_1,\,\lambda_2$, then reflect *perturbation growth increments*, the discrete time analogy of the instantaneous perturbation growth rates of continuous models.

Recall that for continuous time, $\lambda_1$ is the instantaneous rate of change in a perturbation at equilibrium, 
\begin{equation}
 (\#eq:pert1)
  \dot{x}= \lambda_1 x,\quad x_t = x_0e^{\lambda_1 t}
\end{equation}
where $x$ is the perturbation at the equilibrium; if $\lambda_1<0$, the perturbation would decline. For discrete growth, we can think of  $\lambda_1$ as the discrete growth factor of a perturbation at equilibrium, 
\begin{equation}
  (\#eq:pert2)
  x_{t+1}=\lambda_1x_t
\end{equation}
where $x$ is the perturbation. Here, we see that $x$ will decline as long as $0 <\lambda_1 < 1$. Second, if $\lambda_1 < 0$, then $x_{t+1}$ changes sign with each time step, and the perturbation oscillates around 0. That is alright, as the magnitude also decreases with each time step, so the perturbation still declines toward zero. However, if $\lambda < -1$, those oscillations grow and the perturbation oscillates permanently. This is directly analogous to the oscillations, or stable limit cycles, we saw in the discrete logistic growth model. Thus, a criterion for stability in discrete growth models is that for all eigenvalues, $-1 < \lambda < 1$. 

For discrete models, we also need to be a little more concerned about the imaginary part of the eigenvalues, because they contribute to the magnitude of the oscillations and eigenvalues. We therefore add that the magnitude of $\lambda = a + bi$ is  $|\lambda|=\sqrt{a^2 +b^2}$. Thus, the system is stable when $|\lambda|\leq 1$. The magnitude of a complex number is known as its *modulus*.  The moduli (plural of modulus) of the host--parasite model therefore includes the complex plane, were the real part of each eigenvalue is on the $x$-axis, and the imaginary part is on the $y$-axis (Fig. \@ref(fig:mod)). The magnitude or modulus of an eigenvalue is the length of the vector in this plane.

We can illustrate the stability criterion for discrete models in a few ways. The phase plane portrait or time series would illustrate the dynamics directly. However, it is also useful to show the relation between a measure of stability and parameters influencing stability (e.g., Fig. \@ref(fig:LK), \@ref(fig:LK2)?). Since aggregation seems so important in the host--parasitoid model, we can show how stability ($\lambda$) varies with $k$. We can thus investigate *how much* aggregation is required for stability [@Pacala1990]. We would anticipate that stability declines ($|\lambda| \to 1$) as $k$ increases.

Stability of the host--parasitoid model with aggregation (Fig. \@ref(fig:LK2))

We proceed largely as we did for continuous models, first with expressions for, and partial derivatives of, the relevant functions --- for discrete models we use $F(N_t)$, where of $N_{t+1}=F(N_t)$.
```{r discstab}
F.H <- expression(R*H*(1+a*P/k)^-k - H )
F.P <- expression(H - H*(1+a*P/k)^-k - P)
F.H.H <- D(F.H, "H"); F.H.P <- D(F.H, "P")
F.P.H <- D(F.P, "H"); F.P.P <- D(F.P, "P")
```
We next specify a sequence of $k$'s, and for each $k$, find the equilibria, evaluate the Jacobian, and return the eigenvalues of the Jacobian.
```{r}
k <- 10^seq(-1, 1, by=.01)
R <- 3
a <- .005
HPeigs <- sapply(k, function(ki) {
  k <- ki
  P <- k*(R^(1/k) - 1)/a
  H <- P * R/(R-1)
  jac <- matrix( c(eval(F.H.H), eval(F.H.P), eval(F.P.H), eval(F.P.P)), 
                nrow=2, byrow=TRUE )
eigen(jac)[["values"]] 
} )
```

Last, we plot the eigenvalue with the greatest absolute magnitude, and retain the sign of the real part, $\lambda$ \emph{vs.} $k$.

```{r LK2, fig.cap="*Dynamical stability of a discrete host-parasitoid model with aggregation. Region of stability for the rate of change following a small perturbation away from equilibrium. The plus sign '+' is $(0,0)$ and the two small circles are the complex eigenvalues. The length of the vector is the modulus.*"}
modmaxs <- apply( HPeigs, 2, function(lambdas) {i <- which.max(Mod(lambdas))
                          sign(Re(lambdas[i])) * Mod(lambdas[i]) })
plot(k, modmaxs, type='l', ylab=quote("Stability "*(lambda[1])) )
abline(h=-1, lty=3)
```


Most of our eigenvalues are complex numbers.   It is typically important to evaluate the modulus, or magnitude, of the eigenvalues of a Jacobian matrix for a discrete model. Here we graph them in the complex number plane (Fig. \@ref(fig:mod)).

First we set up the unit circle which will define the stability region in the complex number plane.
```{r}
th <- seq(-pi, pi, len=100)
z <- exp(1i*th)
```
We then plot the circle and add the eigenvalues for our smallest $k$; 
```{r}
#The complex number plane
par(pty="s")
plot(z, type="l")
points(0,0, pch=3); points(HPeigs[,100])
arrows(x0=c(0,0), y0=c(0,0), x1=Re(HPeigs[,100]), y1=Im(HPeigs[,100]))
```
The length of the arrows are the moduli, $|\lambda|$.


#### Dynamics of the May host--parasitoid model
Here we simply play with May's model. The following generates a phase plane diagram of the dynamics, although it is not shown. We set the duration of the time series, the model parameters, and create an empty data frame to hold the results.
```{r}
time <- 20; R <- 3; a <- .005; k <- .6
HP2s <- data.frame(Hs = numeric(time), Ps = numeric(time))
```
Next we calculate the equilibrium, and use a nearby point for the initial abundances.
```{r}
P2st <- k*(R^(1/k) - 1)/a; H2st <- P2st * R/(R-1)
P2st;H2st
HP2s[1, ] <- c(1000,500)
```
We then project the dynamics, one generation at a time, for each time step.
```{r}
for(t in 1:(time-1))  HP2s[t+1,]  <- {
  H <- R * HP2s[t,1] * (1+a*HP2s[t,2]/k)^(-k)
  P <- HP2s[t,1] - HP2s[t,1] * (1+a*HP2s[t,2]/k)^(-k)
                               c(H,P) }

```
Last we plot the trajectory, placing a point at the equilibrium, and using arrows to highlight the direction and magnitude of the increment of change per time step.
```{r, NBdynStable}
{
  plot(HP2s[,1], HP2s[,2], type="l", xlab="H", ylab="P")
  points(H2st, P2st)
  arrows(HP2s[-time,1], HP2s[-time,2], HP2s[-1,1], HP2s[-1,2], length=.05)
}
```

## *In Fine*
* Parasitoids are a wildly diverse set of taxa, but largely undescribed.
* We commonly model host-parasitoid in discrete time.
* Hosts grow geometrically, and die when attacked.
* Parasitoids in the next generation are simply the number of attacked hosts.
* Aggregation in attack behavior appears to allow stable coexistence.