---
title: "Beiler et al. 2015"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(igraph)
```

## Exploring Pandora, right here at home

The Na'vi of Pandora know something of how their world works; we are working hard to understand ours. Below our feet, some species cooperate in a sometimes violently antagonistic world - a tangled bank, nature, red in tooth and claw. Suzanne Simard and others have been investigating belowground networks of nutrient flux and information sharing among individuals and between species.


properties assessed

* node = tree
*link = a connection between two trees, 1 or more genets = 1 link 


Network properties

* link density
* degree distribution
* global or network clustering coefficient, global transivity; (no. of closed triplets) / (no. of open + no. closed triplets)
* mean clustering coefficient among nodes in each network
* Betweenness centrality: quantifies the number of times a node acts as a bridge along the shortest path between two other nodes. How important is this node in making connections for other nodes?

Evaluated by testing whether the observed differ from random ER network models based on the same number of nodes and average node degree as empirical MNs


Node properties

* degree
* normalized degree: (d / all possible d)
* node clustering coefficient (local transivity): how close its neighbors are to being a clique themselves. "proportion of links between the vertices within its neighbourhood divided by the number of links that could possibly exist between them" (Wikipedia, clustering coefficient) 


"Network classifications were verified by contrasting empirical networks against undirected random Erdos-Renyi configurations ... having the same number of nodes and mean node degree as the empirical networks, when pairs of nodes are linked at random." 


## Results 

* Vesiculosus more structured than vinicolor
* Vinicolor nested within vesiculosus

No significant relations found between site (soil moisture) and netowrk properties

* network density
* mean node degree, or
* network clustering coefficient 

However, node clustering coefficient and node degree increased with tree dbh (Fig. 4). Seems strange then, that there is a negative relation between node clustering coef and node degree. 


## Data from Beiler et al.

Plot 1
* 26 trees, 258 links (Table 1), node clustering coefficient: 0.84

```{r }
# for Plot 1
V <- 26
E <- 258

gr <- erdos.renyi.game(V, E, type = "gnm")
# graph "order" and "size"
gorder(gr); gsize(gr)

plot(gr, vertex.label= NA, edge.arrow.size=0.02,vertex.size = 0.5, xlab = "Random Network: G(V,E) model")
```

```{r}
# betweenness centrality
bc <- betweenness(gr, v = V(gr), directed = TRUE, weights = NULL,
  nobigint = TRUE, normalized = FALSE)
bcbar <- mean(bc)

# node clustering coefficients (per node)
ncc <- transitivity(gr, type = "local")
# mean node clustering coefficient
mean(ncc) # cf 0.84

# degrees of each node
d <- degree(gr)
# mean degree
mean(d) # cf 18

qplot(d, ncc, geom="jitter")
```

### for all plots

```{r} 
V <- c(26, 13, 31, 27, 27, 41)
E <- c(258, 66, 283, 254, 177, 434)
pV <- c(.846, .5, .875, .846, .778, .889)
dbar <- c(18, 695, 17.15, 17.52, 12.21, 20.19)
nccbar <- c(.84, .63, .84, .85, .83, .8)
plot <- 1:6
df <- data.frame(plot=plot, V=V, E=E, pV=pV, dbar=dbar, nccbar=nccbar)

myStats <- function(ve){
  # ve is a vector of V, E
  V <- ve[1];  E <- ve[2]
  gr <- erdos.renyi.game(V, E, type = "gnm")
  
  bc <- betweenness(gr, directed = FALSE, weights = NULL,
  nobigint = TRUE, normalized = FALSE)
  bcbar <- mean(bc)
  
  #Network clustering coefficient
  gcc <- transitivity(gr)
  
  ncc <- transitivity(gr, type = "local")
# mean node clustering coefficient
 nccbar.o <- mean(ncc) # cf 0.84
# degrees of each node
d <- degree(gr)
# mean degree
dbar.o <- mean(d) # cf 18
return( c(bcbar.op=bcbar, gcc.o=gcc, nccbar.o=nccbar.o, dbar.o=dbar.o))
}
df2 <- cbind(df, t(apply(df[,2:3], 1, myStats)) )
pairs(df2[,5:10])
```
## Fig. 2 but with random graphs
```{r}


ncc1 <- unlist( apply(df[1:3,2:3], 1, function(ve){
   gr <- erdos.renyi.game(ve[1], ve[2], type = "gnm")
   transitivity(gr, type = "local")
 }))
ncc2 <- unlist( apply(df[4:6,2:3], 1, function(ve){
   gr <- erdos.renyi.game(ve[1], ve[2], type = "gnm")
   transitivity(gr, type = "local")
 }))
 
d1 <- unlist( apply(df[1:3,2:3], 1, function(ve){
   gr <- erdos.renyi.game(ve[1], ve[2], type = "gnm")
   degree(gr)
 }))
d2 <- unlist( apply(df[4:6,2:3], 1, function(ve){
   gr <- erdos.renyi.game(ve[1], ve[2], type = "gnm")
   degree(gr)
 }))
dff2 <- data.frame(site = factor(c(rep(1, length(ncc1)), rep(2, length(ncc2)))),
                            ncc =c(ncc1, ncc2), d=c(d1,d2) )
ggplot(dff2, aes(d, ncc, colour=site) ) + geom_jitter()
```

## Fig. 2 but with scale-free graphs
```{r}

exp.out <- 2.1
ncc1 <- unlist( apply(df[1:3,2:3], 1, function(ve){
  gsf <- sample_fitness_pl(ve[1], ve[2], exponent.out= exp.out, exponent.in = -1, loops = FALSE, multiple = FALSE, finite.size.correction = TRUE)
   transitivity(gsf, type = "local")
 }))
ncc2 <- unlist( apply(df[4:6,2:3], 1, function(ve){
  gsf <- sample_fitness_pl(ve[1], ve[2], exponent.out= exp.out, exponent.in = -1, loops = FALSE, multiple = FALSE, finite.size.correction = TRUE)
   transitivity(gsf, type = "local")
 }))
 
d1 <- unlist( apply(df[1:3,2:3], 1, function(ve){
  gsf <- sample_fitness_pl(ve[1], ve[2], exponent.out= exp.out, exponent.in = -1, loops = FALSE, multiple = FALSE, finite.size.correction = TRUE)
   degree(gsf)
 }))
qplot(d1, ncc1)
d2 <- unlist( apply(df[4:6,2:3], 1, function(ve){
   gsf <- sample_fitness_pl(ve[1], ve[2], exponent.out= exp.out, exponent.in = -1, loops = FALSE, multiple = FALSE, finite.size.correction = TRUE)
   degree(gsf)
 }))
dff2 <- data.frame(site = factor(c(rep(1, length(ncc1)), rep(2, length(ncc2)))),
                            ncc =c(ncc1, ncc2), d=c(d1,d2) )
ggplot(dff2, aes(d, ncc, colour=site) ) + geom_jitter()
```

## Fig. 2 but with small-world
```{r}
dm=2
nei=3
p=.2
{
ncc1 <- unlist( apply(df[1:3,2:3], 1, function(ve){
  size <- ceiling((ve[1])^(1/dm))
  gsw <- sample_smallworld(dim=2, size, nei , p )
   transitivity(gsw, type = "local")
 }))
ncc2 <- unlist( apply(df[4:6,2:3], 1, function(ve){
    size <- ceiling((ve[1])^(1/dm))
  gsw <- sample_smallworld(dim=2, size, nei, p)
   transitivity(gsw, type = "local")
 }))
 
d1 <- unlist( apply(df[1:3,2:3], 1, function(ve){
   size <- ceiling((ve[1])^(1/dm))
  gsw <- sample_smallworld(dim=2, size, nei, p)
   degree(gsw)
 }))
qplot(d1, ncc1)
d2 <- unlist( apply(df[4:6,2:3], 1, function(ve){
      size <- ceiling((ve[1])^(1/dm))
  gsw <- sample_smallworld(dim=2, size, nei, p)
   degree(gsw)
 }))
dff2 <- data.frame(site = factor(c(rep(1, length(ncc1)), rep(2, length(ncc2)))),
                            ncc =c(ncc1, ncc2), d=c(d1,d2) )
ggplot(dff2, aes(d, ncc, colour=site) ) + geom_jitter()
}
```