---
title: "Beiler et al. 2015"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(igraph)
```

## Exploring Pandora, right here at home

The Na'vi of Pandora know something of how their world works, and we are working hard to understand ours. Our world is a highly antagonistic world, driven by consumption - a tangled bank, nature, red in tooth and claw. However, all around us and hidden beneath  our feet, many species cooperate to survive and thrive in this otherwise harsh world. This cooperation can happen among species and it can also happen among individuals within a species. Suzanne Simard and others have been investigating belowground networks of nutrient flux and information sharing among individuals and between species.

Most plant species develop mycorrhizae, fungi-root associations which are typically mutualistic interactions. These fungi can often connect individuals of the same species together, creating a network by which one plant can information, nutrients, and carbon with another plant via this fungal highway. However, how these connections are structured is still very poorly known. In an effort to shed some light on this, @Beiler2015 described the topology of Douglas fir trees connected by two species of individual fungal genets, in contrasting environments (\@ref(fig:plot1)).




### Properties assessed

1. Application of networks to mycorrhizae
* node = tree
* link = one or more fungal genets connecting  two trees
* triplet = three connected trees (open triplet is connected by two links, closed, by three)
1. Network properties
* link density (links/species)
* degree distribution (statistical distribution of node degrees)
* global or network clustering coefficient: fraction of triplets that are closed (no. of closed triplets) / (no. of open + no. closed triplets)
* mean lcoal clustering coefficient among nodes in each network
* Betweenness centrality: the number of times a node acts as a bridge along the shortest path between two other nodes. How important is this node in making connections for other nodes?
2. Node properties
* degree (# of links)
* normalized degree: (fraction: d / all possible d)
* node clustering coefficient (local transivity; extent to which its neighbors are are connected)

All these were evaluated by whether the observed properties differed from properties that emerge from random networks with the same number of nodes and average node degree as the empirical mycorrhizal networks. We can generate such a graph using the `igraph` package. 

Plot 1, from a xeric site, had 26 trees and 258 links. Therefore, the average node degree must be $258/26\approx 9.9$. We can create a random graph by simply specify the number of vertices and the number of edges.
```{r plot1. fig.cap="*Example of a random graph using the same number of vertices and links as in plot 1 (Beiler et al. (2015)).*", out.width="70%" }
# node clustering coefficient: 0.84
# for Plot 1
V <- 26
E <- 258

gr <- erdos.renyi.game(V, E, type = "gnm")
# graph "order" and "size"
gorder(gr); gsize(gr)
plot(gr, vertex.label= NA, edge.arrow.size=0.02,vertex.size = 0.5, xlab = "Random Network: G(V,E) model")
```

We can compare the observed average node clustering coefficient ($0.84 \pm 0.37$) to that of the random graph we created above. Note that transivity is node clustering.
```{r}
mean(transitivity(gr, type = "local"))
```
We find that these differ. The observed can then be compared to 1000s of simulated random networks to see how unusual the observed value is. Here we use a toy example with 100 simulations to compare the node clustering coefficient observed in plot 1 to what we would expect in a random network.
```{r}
ncc.sims <- replicate(100, { 
  gr <- erdos.renyi.game(V, E, type = "gnm")
  mean(transitivity(gr, type = "local"))
  } )
qplot(x=ncc.sims, geom="histogram") + 
  geom_vline(xintercept = 0.84, linetype=2)
```

We see that the local node clustering is higher in one real plot than we would expect by chance. Remember that this measures the extent to which a node's neighbors are connected to each other.


```{r}
# betweenness centrality
bc <- betweenness(gr, v = V(gr), directed = TRUE, weights = NULL,
  nobigint = TRUE, normalized = FALSE)
bcbar <- mean(bc)

# node clustering coefficients (per node)
ncc <- transitivity(gr, type = "local")
# mean node clustering coefficient
mean(ncc) # cf 0.84

# degrees of each node
d <- degree(gr)
# mean degree
mean(d) # cf 18

qplot(d, ncc, geom="jitter")
```

Among their results, Beiler et al. found that node clustering coefficient and node degree increased with tree size. It is interesting that in spite of these both increasing with tree size, there is a negative relation between them node clustering coef and node degree (Fig. 2). Beiler et al. point out that this happens in 'scale-free' graphs. 

A *scale-free* graph is one in which the probability of a node having L links follows a power law relation,
$$Pr(L) \sim L^{-z}$$
where, apparently, $2<z<3$. Let's demonstrate that.

First I collect all the data from Beiler et al. (Table 1,2, Beiler et al. (2015)).

```{r} 
# vertices
V <- c(26, 13, 31, 27, 27, 41)
# egdes
E <- c(258, 66, 283, 254, 177, 434)
# proportion of trees colonized
pV <- c(.846, .5, .875, .846, .778, .889)
# mean node degree
mean.deg <- c(18, 695, 17.15, 17.52, 12.21, 20.19)
# mean node clustering coef.
mean.ncc <- c(.84, .63, .84, .85, .83, .8)
plot <- 1:6
df <- data.frame(plot=plot, V=V, E=E, pV=pV, mean.deg=mean.deg, mean.ncc=mean.ncc)
df
```


```{r echo=FALSE, eval=FALSE}
my_ER_Stats <- function(ve){
  # ve is a two-element vector of vertices and edges totals
  
  V <- ve[1];  E <- ve[2]
  gr <- erdos.renyi.game(V, E, type = "gnm")
  
  # betweenness centrality
  bc <- igraph::betweenness(gr, directed = FALSE, weights = NULL,
  nobigint = TRUE, normalized = FALSE)
  bcbar <- mean(bc)
  
  # Network clustering coefficient
  gcc <- transitivity(gr) # global
  
  ncc <- transitivity(gr, type = "local")
# mean node clustering coefficient
 nccbar.o <- mean(ncc) # cf 0.84
# degrees of each node
d <- igraph::degree(gr)
# mean degree
dbar.o <- mean(d) # cf 18
return( c(bcbar.op=bcbar, gcc.o=gcc, nccbar.o=nccbar.o, dbar.o=dbar.o))
}
```

Next I make a function that generates a random graph from the data in Beiler et al. and calculates the propoerties I need.

For scale free graphs, I will set the power exponent to something between 2 and 3.
```{r}
exp.out <- 2.1
```

The following somewhat complicated looking code does the following for each plot in each site

* generates a scale-free random graph
* calculates local node clustering coefficient (transivity) for each resulting "tree",
* calculates the degree for each "tree",
* records the site of the plot.

It then binds together the data collected for each plot at each site, and binds together the data freom both sites.
```{r}
# Site 1 random graph and propoerties of "trees"
ncc1 <- apply(df[1:3,2:3], 1, function(ve){
  gsf <- sample_fitness_pl(ve[1], ve[2], exponent.out= exp.out,
                           exponent.in = -1, loops = FALSE,
                           multiple = FALSE, 
                           finite.size.correction = TRUE)
  ncc <- transitivity(gsf, type = "local")
  deg <- igraph::degree(gsf)
  m <- matrix(c(ncc, deg, site=rep(2, length(deg))), ncol=3)
  m
 })

# Site 2 random graph and propoerties of "trees"
ncc2 <- apply(df[4:6,2:3], 1, function(ve){
  gsf <- sample_fitness_pl(ve[1], ve[2], exponent.out= exp.out,
                           exponent.in = -1, loops = FALSE,
                           multiple = FALSE, 
                           finite.size.correction = TRUE)
  ncc <- transitivity(gsf, type = "local")
  deg <- igraph::degree(gsf)
  m <- matrix(c(ncc, deg, site=rep(2, length(deg))), ncol=3)
  m
 })

# bind together plots within each site
site1 <- do.call(rbind, ncc1)
site2 <- do.call(rbind, ncc2)

# bind together both sites and label columns
df.sf <- rbind.data.frame(site1, site2)
names(df.sf) <- c("ncc", "deg", "site")

# plot the relation between node clustering coefficient
# and node degree
ggplot(dff2, aes(d, ncc, colour=site) ) + geom_point() 
```

The result is one example of what we would observe in a scale-free network. Many observed networks appear to be scale-free, at least at first glance [@Barabasi:2009aa]. They may be a useful approximation of networks where node properties might be self-reinforcing, that is the better-connected nodes tend to become ever better connected. 
