---
title: "Untitled"
output: html_document
---
### Proportional change
If all we are interested in is the proportional change with each time step, that is simply 
$$\frac{N_{t+1}}{N_t}$$
where $t$ is a particular time step, suych as two weeks. For our flies, that is simply
```{r}
N[2:4] / N[1:3]
```
showing that the population doubles every two weeks. When we treat this proportional cahnge as a constant, we typically symbolize it as $\frac{N_{i+1}}{N_i} = \lambda$, and refer to it as the *finite rate of increase* of a population.

If we want to project the population size into the future, we can use $\lambda$, where 
$$N_{t+1} = N_t \lambda$$
$$N_{t+2} = N_{t+1} \lambda = \left(N_{t}\lambda\right) \lambda $$
$$N_{t+3} = N_{t+2} \lambda = \left(N_{t}\lambda\right) \lambda \lambda$$
From this we can infer this rule:
$$N_{t} = N_0 \lambda^t$$
where $N_0$ is the initial population size, and $t$ is however many time steps we want to project into the future. We refer to this as the *projection* form of geometric growth.

When $\lambda = 1$ the population is neither growing nor shrinking. When $\lambda > 1$, the population is growing, and hen $\lambda < 1$, the population is shrinking. Finally, $\lambda$ can never be less than zero.




### Growth rates
The *population growth rate* the *change in population size per unit time*. In this case we can say
$$\frac{\Delta N}{\Delta t} = \frac{N_{t+\Delta t}-N_t}{\Delta t}$$
where $t$ is any point in time, and $\Delta t$ is interval between successive time points. In our case, if $t=0$, then $N_0 = 2$, and $N_{0+2} = 4$. Therefore, we have 
$$\frac{\Delta N}{\Delta t} = \frac{4-2}{2} = 1$$ flies per week, at least in the first week. Let's do that calculation for each time period.

The *per capita growth rate* is the population growth rate divided by $N$, or
$$\frac{\Delta N}{\Delta t N_t} = \frac{N_{t+\Delta t}-N_t}{\Delta t} \frac{1}{N_t}$$
```{r}
# change in population sizes
DeltaN <- N[2:4] - N[1:3]
Deltat <- t[2:4] - t[1:3]
DN.Dt <- DeltaN / Deltat
DN.DtN <- DN.Dt / N[1:3]
cbind(Deltat, DeltaN, DN.Dt, DN.DtN)
```
From these columns of calculations, we see that (i) the time intervals are constant (two weeks apart), (ii) the change in population size is getting progressively bigger, (iii) the population growth rate is increasing, and (iv) the per capita growth rate is constant. 

If we wish to project the size of the population into the future, we can multiply the size of the population times the per capita growth rate and add that to the population. Let the per capita growth rate be 
$$r_d=\frac{\Delta N}{ \Delta t N_t}$$.

To project the population, we simply let

$$N_{t + \Delta t} = N_t + N_t r_ = N_t(1+r_d)$$
$$N_{t + 2\Delta t} =  N_{t+\Delta t}(1+r_d) = N_t(1+r_d)^2$$

Now we see the relation between the per capita growth rate, the population growth, and the finite rate of increase.

### Fruit flies with continuous overlapping generations

In the reality that is my aseasonal kitchen, individual fruit flies are having sex and reproducing on their own schedules - they breed continuously and the cohorts are not synchronous. For populations like that, we need to describe instantaneous growth rates, where $\Delta t$ i no longer a fixed period of time, but is the rate at a given instant. For this we need differential colculus, where the differential rate equation is
$$\frac{dN}{dt} = rN$$
Note the similarity to and difference from $\lambda$. When $ r = 0$ the population is neither growing nor shrinking. When $r > 0$, the population is growing, and when $r < 0$, the population is shrinking. 

To project a population, we integrate $rN$ with respect to time.
$$dN/N = r dt$$
$$\int_{N_0}^{N_t} \frac{1}{N}dN = \int_{0}^{t}rdt$$


```{r diagram, eval=FALSE, echo=FALSE}
library(diagram)
par(mar=c(1,1,1,1))
names <- c("A", "F", "B")
f <- matrix(0, nc=3, nr=3)

f[2,1] <- "I"
f[3,2] <- "E~D"
f[2,2] <- "B"
arr.hd <- matrix(c(0,1,0, 0,1,1, 0,0,0), nr=3)
pp <- plotmat(f,pos=c(3),name=names,lwd=1,box.lwd=2,cex.txt=0.8,
             box.type="circle", box.size=0.05, box.lcol=c(0,1,0),
             shadow.size=0, arr.pos=arr.hd,
             self.shiftx=0, self.shifty=-.05, self.arrpos=.5)
```
