# Diversity

## Background
```{r BSsuc,fig=TRUE, fig.width=2.5, fig.height=2.5, echo=FALSE, results='hide', message=FALSE, fig.cap="Empirical rank--abundance distributions of successional plant communities (old-fields) within the temperate deciduous forest biome of North America. ``Year'' indicates the time since abandonment from agriculture.  Data from the Buell-Small succession study (http://www.ecostudies.org/bss/)"}
bsdiv <- read.csv('C3rank.csv')
bs <- melt(bsdiv, id="Rank", variable.name="Year", value.name="Percent_cover")
str(bs)
ggplot(bs, aes(Rank, Percent_cover, linetype=Year, colour=Year)) + geom_line() + scale_y_log10()
```

It seems easy, or at least tractable, to compare the abundance of a
single species in two samples. In this chapter, we introduce concepts
that ecologists use to compare entire communities in two samples. We
focus on two quantities: *species composition*, and
*diversity*. We also discuss several issues related to this,
including species--abundance distributions,  ecological neutral
theory, diversity partitioning, and species--area relations. Several packages in R include functions for dealing specifically with these topics. 
Please see the ``Environmetrics'' link within the ``Task Views'' link at any CRAN website for downloading R packages. Perhaps the most comprehensive (including both diversity and composition) is the `vegan` package, but many others include important features as well.

## Neutral theory

Species area immigration curves
```{r}
d <- .4
i <- 1:26
g <- d*(1-d)^(i-1)
geodist <- g/sum(g)
comm <- sample(letters, 1e4, replace=TRUE, prob=geodist)
plot( table(comm) )
```

Gamma function
```{r}

gamma(3) # 2*1
gamma(4) # 3*2*1
gamma(3.5) # something in between. :)
```