# Mutualisms

## Background

Mutualisms occur in a wide variety of different species. Mutualisms form the
foundations of ecosystems including coral reefs (coral and
zooxanthellae), grasslands, and forests (plants and mycorrhizal
fungi, pollinators, and dispersers). 

The natural history of mutualisms are wildly diverse and complex, and 
and @Boucherecologymutualism1982 lists types of mutualisms:

* Energetic
* Nutritional
* Defense
* Transport
* Obligate \emph{vs.} facultative
* Direct \emph{vs.} indirect
* Symbiotic \emph{vs.} independent

 Perhaps the most common of these is
*indirect mutualism* (Fig. \ref{f:mut3}). In this case, negative direct interactions
sum to yield net positive indirect interactions, as in the classic
case where ``the enemy of my enemy is my friend.'' For instance,
consider a three plant species that all compete for the same limiting
resource. When species A suppresses species B, it is indirectly
helping species C [@Miller1994]. 

## Lotka-Volterra mutualism
Lotka-Volterra mutualism is very straightforward, following the samed template as competition. All that we have to do is change the sign associated with the other species. As usual, $\alpha_{ij}$ is the per capita effect of species $j$ on species $i$.

\begin{align}
\frac{dN_1}{dt} &= r_1N_1\left( 1 - \alpha_{11}N_1 + \alpha_{12}N_2\right)\\
\frac{dN_2}{dt} &= r_2N_2 \left(1 + \alpha_{21}N_1 - \alpha_{22}N_2\right)
\end{align}

Here we will focus on *direct mutualisms*, and in this chapter, we'll focus on two approaches to models, Lotka-Volterra models and consumer-resource models.

In the above equations we see the sign of the intraspecific terms $\alpha_{ii}N_i$ is negative, while the sign of the interspecific terms $\alpha_{ij}N_j$.

The zero net growth isoclines are found the same way we found those for interspecific competition, by solving $\frac{dN_1}{dt}$ (for zero). We will graph these, so we solve each in terms of species 1.

The ZNGI for species 2 is
\begin{align*}
0 &= 1 + \alpha_{21}N_1 - \alpha_{22}N_2\\
N_2 &= 1/\alpha_{22} + \alpha_{21}N_1
\begin{align*}
Thus when $N_1=0$, $N_2 = 1/\alpha_22} = K_2$.

The ZNGI for species 1, in terms of species 1, is
\begin{align*}
0 &= 1 - \alpha_{11}N_1 + \alpha_{12}N_2\\
N_2 &= \alpha_{11}/\alpha_{12}N_1 - 1/\alpha_{12}
\begin{align*}

```{r}
alpha11 <- alpha22 <- 0.2
alpha12 <- alpha21 <- 0.1

plot(0:5, 0:5, type="n")
abline(1/alpha22, alpha21)
abline(1/alpha12, alpha11/alpha12, lty=2)

```

## Consumer-resource mutualism
Sometimes mutualists share resources, as with mycorrhizal
associations, where species share a resource or a product that they
can acquire or manufacture in surplus and trade it for a resource or product that is
limiting. This is a bi-drectional mutualism (Fig. \ref{f:mut3})
because the flow of \emph{resources} are bidirectional: both species provide a resource to the other [@Holland2010]. 
```{r figs, echo=FALSE}
# a digraph is a directed graph
# the model is written in the "dot" language
bimut <-
"digraph{
  rankdir = LR;
node[shape=box];
// oval is the default node shape
  M1 -> M2[label='R'];
M2 -> M1[label='R'];
}"

#grViz(bimut)

umut <-
"digraph{
  rankdir = LR;
node[shape=box];
// oval is the default node shape
  M1 -> M2[label='R'];
M2 -> M1[label='S', style=dashed]
}"

imut <-
"digraph{
  rankdir = TB;
node[shape=box];
{rank=max; R1; R2;}
{rank=min; M1; M2;}
// oval is the default node shape
  M1 -> M2[label='R'];
M2 -> M1[label='S', style=dashed]
R1 -> M1; M1 -> R1[arrowhead='odot'];
R2 -> M2; M2 -> R2[arrowhead='odot'];
R1 -> R2[arrowhead='odot']; R2 -> R1[arrowhead='odot'];
}"

grViz(imut)

grViz(bimut) %>%
  export_svg %>% charToRaw %>% rsvg_pdf("bimut.pdf")

grViz(umut) %>%
  export_svg %>% charToRaw %>% rsvg_pdf("umut.pdf")

grViz(imut) %>%
  export_svg %>% charToRaw %>% rsvg_pdf("imut.pdf")
```

\begin{figure}
\begin{center}
\hfill
\includegraphics[width=.4\linewidth]{bimut} 
\hfill
\includegraphics[width=.4\linewidth]{umut}
\hfill 
\\
\includegraphics[width=.4\linewidth]{imut}
\caption{Bidirection, unidirectional, and indirect mutualisms. Normal
  arrowhead is a positive effect, whereas an open dot arrowhead is a
  negative effect. A dashed line indicates a service, and a solid line
  indicates a consumeable resource. \label{fig:mut3}}
\end{center}
\end{figure}

In other cases, mutualists may share services, as with moray
eel-grouper foraging. In this particular case, grouper and the giant
moray eel have different foraging strategies, with the eel going into
reef crevices and the grouper waiting outside to pick off prey that
escape the eel [@Bshary2006]. Sometimes these mutualisms are
called *by-product* mutualisms, in cases where there is no
evidence of coordinated behavior, but nonetheless a mutualism arises via the independent activities of each species.

In the following, we explore the type of mutualism in which one
species provides a resource and the other species provides a
service (Fig. \ref{fig:mut3}). This is a uni-directional mutualism because
the flow of \emph{resources} is only unidrectional, from one species to the other [@Holland2010].

A good example of a uni-directional of mutualism is seed dispersal via
a frugivore that disperses seeds. A plant species provides a 
fruit that is nutritious for an animal. A primary benefit for the
plant is that seeds are carried away from the parent plant. This may
help seeds escape enemies that accumulate near parent plants
(Connell-Janzen hyptothesis). In addition, it provides a bet hedging
opportunity wherein seeds are dispersed to a variety of habitats any
one of which may be better for survival than the current location. It
is even possible that dispersal is *directed* to preferred
habitat that is consistently better than under the parent plant due to
better resource levels or environmental conditions
[@Wenny1998]. The animal benefits by consuming the fleshy part of
the fruit surrounding the seed. An interesting case arises when seed
predators collect seeds but do not kill them before having buried,
cached, or moved seeds. 