---
title: "9.3"
author: "MHHS"
date: "12/31/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Stability of host--parasitoid dynamics
As with differential equation models, we can analyze the stability in the immediate vicinity of equilibria in discrete time models using eigenanalysis of the Jacobian matrix. Although analogous, there are some important differences  between continuous and discrete models.

* The Jacobian matrix is comprised of partial derivatives of growth increments (e.g., $\Delta N = N_{t+1}-N_t$), rather than of time derivatives (e.g., $\dot{N}$).
* Interpretation of eigenvalues of the Jacobian reflects the difference between finite rate of increase ($\lambda$) and the intrinsic rate of increase ($r$); e.g., for discrete models, $\lambda = 1$ indicates zero growth.
* Populations with discrete generations can grow too quickly to be stable (e.g., chaos in discrete logistic growth).


In the discrete time case, the Jacobian matrix is the set of partial derivatives of the discrete growth increment rather than of the time derivatives used for continuous growth^[Note $dN/dt$ is also an increment --- it is the increment $\Delta N$ as $\Delta t \to 0$, whereas in the discrete case, $\Delta t = 1$.]. The growth increments are the increments of change over an entire time step, or $\Delta H=H_{t+1} - H_t$ and $\Delta P=P_{t+1} - P_t$. Taking those increments from \@ref(eq:may14a), \@ref(eq:may14b), the Jacobian matrix of partial differential equations is 
\begin{align}
  (\#eq:jachp)
\left(
\begin {array}{cc}
\frac{\partial \Delta H}{\partial H}&\frac{\partial \Delta H}{\partial P}\\
\frac{\partial \Delta P}{\partial H}&\frac{\partial \Delta P}{\partial P}
\end {array}
\right) =
\left(
\begin {array}{cc}
R\left(1+aP\right)^{-k}-1\quad &-akHR\left(1+aP\right)^{-\left(k+1\right)} \\
1-\left(1+aP\right)^{-k}& akH\left(1+aP\right)^{-\left(k+1\right)}
\end {array}
\right) 
\end{align}

To analyze this at the equilibrium, we substitute $H^*,\,P^*$ for $H,\,P$, and perform eigenanalysis. The resulting eigenvalues, $\lambda_1,\,\lambda_2$, then reflect *perturbation growth increments*, the discrete time analogy of the instantaneous perturbation growth rates of continuous models.

Recall that for continuous time, $\lambda_1$ is the instantaneous rate of change in a perturbation at equilibrium, 
\begin{equation}
 (\#eq:pert1)
  \dot{x}= \lambda_1 x,\quad x_t = x_0e^{\lambda_1 t}
\end{equation}
where $x$ is the perturbation at the equilibrium; if $\lambda_1<0$, the perturbation would decline. For discrete growth, we can think of  $\lambda_1$ as the discrete growth factor of a perturbation at equilibrium, 
\begin{equation}
  (\#eq:pert2)
  x_{t+1}=\lambda_1x_t
\end{equation}
where $x$ is the perturbation. Here, we see that $x$ will decline as long as $0 <\lambda_1 < 1$. Second, if $\lambda_1 < 0$, then $x_{t+1}$ changes sign with each time step, and the perturbation oscillates around 0. That is alright, as the magnitude also decreases with each time step, so the perturbation still declines toward zero. However, if $\lambda < -1$, those oscillations grow and the perturbation oscillates permanently. This is directly analogous to the oscillations, or stable limit cycles, we saw in the discrete logistic growth model. Thus, a criterion for stability in discrete growth models is that for all eigenvalues, $-1 < \lambda < 1$. 

For discrete models, we also need to be a little more concerned about the imaginary part of the eigenvalues, because they contribute to the magnitude of the oscillations and eigenvalues. We therefore add that the magnitude of $\lambda = a + bi$ is  $|\lambda|=\sqrt{a^2 +b^2}$. Thus, the system is stable when $|\lambda|\leq 1$. The magnitude of a complex number is known as its *modulus*.  The moduli (plural of modulus) of the host--parasite model therefore includes the complex plane, were the real part of each eigenvalue is on the $x$-axis, and the imaginary part is on the $y$-axis (Fig. \@ref(fig:mod)). The magnitude or modulus of an eigenvalue is the length of the vector in this plane. 

We can illustrate the stability criterion for discrete models in a few ways. The phase plane portrait or time series would illustrate the dynamics directly. However, it is also useful to show the relation between a measure of stability and parameters influencing stability. Since aggregation seems so important in the host--parasitoid model, we can show how stability ($\lambda$) varies with $k$. We can thus investigate *how much* aggregation is required for stability [@Pacala1990]. We would anticipate that stability declines ($|\lambda| \to 1$) as $k$ increases.

We proceed largely as we did for continuous models, first with expressions for, and partial derivatives of, the relevant functions --- for discrete models we use $F(N_t)$, where of $N_{t+1}=F(N_t)$.
```{r discstab}
F.H <- expression(R*H*(1+a*P/k)^-k - H )
F.P <- expression(H - H*(1+a*P/k)^-k - P)
F.H.H <- D(F.H, "H"); F.H.P <- D(F.H, "P")
F.P.H <- D(F.P, "H"); F.P.P <- D(F.P, "P")
```
We next specify a sequence of $k$'s, and for each $k$, find the equilibria, evaluate the Jacobian, and return the eigenvalues of the Jacobian.
```{r}
k <- 10^seq(-1, 1, by=.01)
R <- 3
a <- .005
HPeigs <- sapply(k, function(ki) {
  k <- ki
  P <- k*(R^(1/k) - 1)/a
  H <- P * R/(R-1)
  jac <- matrix( c(eval(F.H.H), eval(F.H.P), eval(F.P.H), eval(F.P.P)), 
                nrow=2, byrow=TRUE )
eigen(jac)[["values"]] 
} )
```

Last, we plot the eigenvalue with the greatest absolute magnitude, and retain the sign of the real part, $\lambda$ \emph{vs.} $k$.

```{r LK, fig.cap="*Dynamical stability of a discrete host-parasitoid model with aggregation. Region of stability for the rate of change following a small perturbation away from equilibrium. The plus sign '+' is $(0,0)$ and the two small circles are the complex eigenvalues. The length of the vector is the modulus.*"}
modmaxs <- apply( HPeigs, 2, function(lambdas) {
  i <- which.max(Mod(lambdas))
  sign(Re(lambdas[i])) * Mod(lambdas[i]) })

plot(k, modmaxs, type='l', 
     ylab=quote("Stability "*(lambda[1])) )
abline(h=-1, lty=3)
```


Most of our eigenvalues are complex numbers.   It is typically important to evaluate the modulus, or magnitude, of the eigenvalues of a Jacobian matrix for a discrete model. Here we graph them in the complex number plane (Fig. \@ref(fig:mod)).

First we set up the unit circle which will define the stability region in the complex number plane.
```{r}
th <- seq(-pi, pi, len=100)
z <- exp(1i*th)
```
We then plot the circle and add the eigenvalues for our smallest $k$; 
```{r LK2, fig.cap="*The complex number plane; the length of the arrows are the moduli, lambda.*"}
par(pty="s")
plot(z, type="l")
points(0,0, pch=3); points(HPeigs[,100])
arrows(x0=c(0,0), y0=c(0,0), x1=Re(HPeigs[,100]), y1=Im(HPeigs[,100]))
```
The length of the arrows are the moduli, $|\lambda|$.

